Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-19 doc/thesis/intro.tex
Finished draft intro!

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. \defn{Inter-host} phylogenies relate one representative viral
genotype from each host in a population. In this work, when we use the phrase
``viral phylogeny'', we are referring to an inter-host phylogeny unless
otherwise stated. Phylogenetic inference is a complex topic which we will not
discuss here (see \eg \autocite{nei2000molecular} for a full review). We assume
here that the viral phylogeny is known, a common assumption in the development
and application of phylodynamic methods.

The crux of phylodynamics is the fact that the epidemiological processes
recorded in transmission trees, and the evolutionary processes recorded in
viral phylogenies, occur on similar time scales for RNA
viruses~\autocite{drummond2003measurably}. Phylodynamic methods have been used
to investigate epidemiological parameters such as transmission rate, recovery
rate, and basic reproductive number~\autocite{pybus2009evolutionary,
volz2013viral}. The majority of phylodynamic studies attempt to infer the
parameters of an epidemiological model for which the likelihood of an observed
phylogeny can be calculated. Most often, this is some variation of the
birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
coalescent~\autocite{kingman1982coalescent, volz2012complex} models.

Due to the relationship between the aforementioned processes, there is a close
relationship between viral phylogenies and transmission trees. In particular,
the transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Due to these relationships, the topology of the viral phylogeny is
often used as a proxy for the topology of the transmission tree. However, there
are several complications and caveats which must be kept in mind when
estimating the transmission tree in this manner.

First are the issues of rooting and time-scaling. Modern likelihood-based
methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. It is generally assumed that sampling a
virus from the individual also corresponds to their removal from the
transmission tree, so the positions of the tips in time are fixed. Therefore,
to transform a viral phylogeny into an estimated transmission tree, we must
find a root and an assignment of branch lengths such that the tips are placed
at their proscribed times. In addition, the branch lengths should be chosen
such that the variation in \defn{evolutionary rate}, which is the ratio of a
branch's length in genetic distance to its length in calendar time, is low
across the tree. While there is some variation among hosts, due to
immunological and other factors, we generally expect to observe globally
similar evolutionary rates. Methods for time-scaling a phylogeny include
root-to-tip regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Both of these methods can be used to root the
tree, by simply trying all possible root positions and choosing the one which
minimizes a loss function ($1-R^2$, or root-mean-square error, for root-to-tip
regression; sum of squared errors for least-square dating). Alternatively, the
tree may be rooted separately with an outgroup~\autocite{li1988rates} before
time-scaling.

A second, perhaps more insidious problem is the fact that the correspondence
between the topologies of the viral phylogeny and transmission tree is not
necessarily exact. Due to intra-host diversity, the viral strain which is
transmitted may have split from another lineage within the donor long before
the transmission event occurred. Hence, the branching point in the viral
phylogeny may be much earlier than that in the transmission tree. Another
possibility is that one host transmitted to two or more recipients, but the
lineages they each received originated within the donor host in a different
order than that in which the transmissions occurred. In this case, the topology
of the transmission tree and the viral phylogeny will be mismatched. Although
phylodynamics is quite new, these phenomena have been studied in evolutionary
biology for some time. Viral phylogenies are a specific version of a more
general class of trees called \defn{gene trees}, which represent the
evolutionary history of a section of genetic material. Transmission trees, on
the other hand, are highly analogous to \defn{species trees}, whose tips are
species and internal nodes are common ancestors. This analogy derives from the
functional similarity between transmission and allopatric speciation. Hence,
the potential discordance between transmission trees and viral phylogenies is
the similar to that between gene and species trees, which is called
\defn{incomplete lineage sorting}. In practice, this discordance has not proven
an insurmountable problem: for example, \textcite{leitner1996accurate} were
able to accurately recover a known transmission tree using a viral phylogeny.

What we have just discussed is a two-step procedure for estimating the
transmission tree. First, a viral phylogeny is constructed from genetic
sequence data, and then it is rooted and time-scaled into a transmission tree.
This approach is straightforward, frequently used, and has the advantage of
leveraging tried-and-true tools for phylogenetic inference. However, it also
has drawbacks, perhaps the most obvious being the multiplication of errors
produced by the separate steps. One commonly used alternative method is to
directly estimate a time-scaled phylogeny by simultaneously inferring the tree
topology, its root and branch lengths, and the parameters of a \defn{molecular
clock} model. A molecular clock is a hypothesis about the evolutionary rates
along the branches of the tree, such as that they are all equal (a \defn{strict
clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
clock}). This inference is usually done in a Bayesian framework using
\gls{MCMC}, so that prior information (including the tip dates) can be included
in the analysis, and the so-called nuisance parameters of the molecular clock
model can be marginalized out. Software packages for performing these analyses
include BEAST~\autocite{bouckaert2014beast} and
MrBayes~\autocite{ronquist2012mrbayes}.

Several other authors have developed methods tailor-made for inferring
transmission trees. \textcite{didelot2014bayesian} develop a Bayesian version
of the two-step approach which allows transmissions to occur anywhere along the
branches of a transmission tree, rather than being constrained to the branching
points in the viral phylogeny. The method requires sampling of every infected
individual, although the authors indicate that it could be extended to relax
this assumption. \textcite{cottam2008integrating} describe a likelihood-based
method which enumerates all transmission trees consistent with an established
phylogeny, assigning each a likelihood based on other epidemiological data. 
This approach is novel in its integration of data from multiple sources,
however because it enumerates a large portion of the tree space, it is unlikely
to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint 
likelihood function integrating temporal, geographic, and genetic observations, 
and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
likelihood function. Their approach can handle missing data and produces high
resolution transmission trees when multiple types of data are available. A
different approach is undertaken by \textcite{jombart2011reconstructing}, who
describe a method to build transmission trees directly from sequence data,
contingent on the common ancestors also being sampled. This makes the method
attractive for slow-evolving pathogens, but less practical for viral outbreaks
where samples from common ancestors are unlikely to be available.

@@

The aim of viral phylodynamics is to glean some kind of knowledge, about the
epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
most often an estimated transmission tree~\autocite{pybus2009evolutionary,
volz2013viral}. What is informative about a phylogeny, beyond the demographic
characteristics of the individuals it relates, is its \defn{shape}. The shape
of a phylogeny has two components: the topology, and the distribution of branch
lengths~\autocite{mooers1997inferring}. Methods of quantifying tree shape fall
into two categories: summary statistics, and pairwise measures.

Summary statistics assign a numeric value to each individual tree. One of the
most widely used is Sackin's index~\autocite{shao1990tree}, which measures the
imbalance or asymmetry in a rooted tree. For the $i$th tip of the tree, we
define $N_i$ to be the number of branches between that tip and the root. The
unnormalized Sackin's index is defined as the sum of all $N_i$. It is called
unnormalized because it does not account for the number of tips in the tree.
Among two trees having the same number of tips, the least-balanced tree will
have the highest Sackin's index. However, among two equally balanced trees, the
larger tree will have a higher Sackin's index. This makes it challenging to
compare balances among trees of different sizes. To correct this,
\textcite{kirkpatrick1993searching} derive the expected value of Sackin' index
under the Yule model~\autocite{yule1925mathematical}. Dividing by this expected
value normalizes Sackin's index, so that it can be used to compare trees of
different sizes.

Rather than assigning numbers to individual trees, pairwise measures associate
a numeric value to each \textit{pair} of trees, indicating how different the
trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion. One such distance measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the difference between the two plots is integrated between 0 and 1. In
the context of infectious diseases, the \gls{ltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = <\varphi(x), \varphi(x')>,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function $K$ is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. The feature space into which trees are mapped is the space of all
possible \defn{subset trees}, which are subtrees that do not necessarily extend
all the way to the tips. The subset-tree kernel was originally developed for
comparing parse trees in natural language processing~\autocite{collins2002new}
and did not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

@@

Several properties of scale-free networks are relevant in epidemiology.  The
high-degree hub nodes are known as
\defn{superspreaders}~\autocite{kemper1980identification}, which have been
postulated to contribute in varying degree to the spread of diseases such as
\gls{HIV}~\autocite{stadler2013uncovering} and
\gls{SARS}~\autocite{shen2004superspreading}. Scale-free networks have no
epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
arbitrarily low transmissibility can persist at low levels indefinitely. This
is in contrast with homogeneously mixed populations, in which transmissibility
below the epidemic threshold would result in exponential decay in the number of
infected individuals and eventual extinction of the pathogen (Anderson \& May,
I think).

@@

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2010contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by
\textcite{volz2009phylodynamics}. The concordance between the two skyrides, as
well as the relationship between the skyride and prevalence curve, was
qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and found a moderate
correspondence between the two.

@@

\Gls{SMC} is a statistical inference method which samples from a sequence of
probability distributions in a fixed order~\autocite{del2006sequential}, based
on a technique called \gls{SIS}.

@@

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more precise definition can be found in
\textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
functions defined in \cref{subsec:treeshape}, other than sharing a name (the
word ``kernel'' is ubiquitous in mathematics).

\subsection{Sequential importance sampling}
\label{subsec:sis}

The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
statistics of distributions which are known only up to a normalizing constant,
and therefore cannot be sampled from directly. That is, if $\pi$ is such a
distribution and $f$ is any real-valued function, \gls{IS} is concerned with
estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We can
express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
we need to obtain an estimate of $\pi(f)$. Although this is a simple and
elegant approach, the drawback is that the variance of the estimate is
proportional to the variance of the importance weights~\autocite{liu2008monte},
which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
the practical use of \gls{IS} on its own is limited, since it depends on
finding an importance distribution similar to $\pi$, which we usually know very
little about \textit{a priori}.

\gls{SIS} applies \gls{IS} to the problem of sampling from a distribution $\pi$ 
on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The
objective is to build up an importance distribution $\eta$ for $\pi$
sequentially. By the general product rule, $\pi(\vec{x})$ can be decomposed as
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ can be written recursively as
\begin{align}
  \label{eq:sisw}
  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
\end{align}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. This choice is made on a
problem-specific basis, taking any available information about $\pi(x_i \mid
\vec{x_{i-1}})$ into account (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).  One potential
choice for $\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid
\vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
prior distribution may be used. The exact form of $\eta(x_i, \vec{x_{i-1}})$
which minimizes the variance of the weights is called the \defn{optimal
kernel}~\autocite{cappe2007overview}, the name deriving from the fact that
$k(x_i, \vec{x_{i-1}}) = \eta(x_i, \vec{x_{i-1}})$ is a Markov kernel. In some
applications, it is possible to approximate the optimal kernel or even compute
it explicitly.

@@

\subsection{Sequential Monte Carlo samplers}
\label{subsec:smcsamp}

@@

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $\vec{x} = \set{x_1, \ldots, x_n}$,
and outcomes, $\vec{y} = \set{y_1, \ldots, y_n}$. A \defn{parametric} model
defines a family of possible relationships between data and outcomes, indexed
by one or more numeric parameters $\theta$. A \defn{statistical} model
describes the relationship between data and outcomes in terms of probabilities.
Statistical models define, either explicitly or implicitly, the probability of
observing $\vec{y}$ given $\vec{x}$ and, if the model is parametric, $\theta$.
In this context, the observed outcomes are taken to be realizations of random
variables $\vec{Y} = \set{Y_1, \ldots, Y_n}$. Note that it is entirely possible
to have no data $\vec{x}$, only observed outcomes $\vec{y}$. In this case, a
model would describe the process by which $\vec{y}$ is generated.

>>>>

The basis of phylodynamics is the fact that, for RNA viruses, epidemiological
and evolutionary processes occur on similar time
scales~\autocite{drummond2003measurably}. In fact, these two processes
interact, such that it is possible to detect the influence of host epidemiology
on the evolutionary history of the virus as recorded in an \defn{inter-host
viral phylogeny}. Phylodynamic methods aim to detect and quantify the
signatures of epidemiological processes in these
phylogenies~\autocite{pybus2009evolutionary, volz2013viral}, which relate one
representative viral genotype from each host in an infected population. These
methods have been used to investigate parameters such as transmission rate,
recovery rate, and basic reproductive number~\autocite{pybus2009evolutionary,
volz2013viral}. The majority of phylodynamic studies attempt to infer the
parameters of an epidemiological model for which the likelihood of an observed
phylogeny can be calculated. Most often, this is some variation of the
birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
coalescent~\autocite{kingman1982coalescent, volz2012complex} models. These
methods generally assume the viral phylogeny is known, and we assume the same
in this work (see \eg \autocite{nei2000molecular} for a review of phylogenetic
inference).

Due to the relationship between the aforementioned processes, there is a degree
of correspondence between viral phylogenies and transmission
trees~\autocite{ypma2013relating, kenah2015algorithms, kenah2016molecular}. In
particular, the transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
as a proxy for the topology of the transmission tree. Modern likelihood-based
methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. Therefore, estimating a transmission
tree from a viral phylogeny requires the phylogeny to be rooted and
time-scaled. Methods for performing this process include root-to-tip
regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
with an outgroup~\autocite{li1988rates} before time-scaling.

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is not necessarily exact~\autocite{ypma2013relating}. Due to intra-host
diversity, the viral strain which is transmitted may have split from another
lineage within the donor long before the transmission event occurred. Hence,
the branching point in the viral phylogeny may be much earlier than that in the
transmission tree. Another possibility is that one host transmitted to two or
more recipients, but the lineages they each received originated within the
donor host in a different order than that in which the transmissions occurred.
In this case, the topology of the transmission tree and the viral phylogeny
will be mismatched. In practice, this discordance has not proven an
insurmountable problem: for example, \textcite{leitner1996accurate} were able
to accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, didelot2014bayesian}.

%Although phylodynamics is quite new, these phenomena have been studied in
%evolutionary biology for some time. Viral phylogenies are a specific version of
%a more general class of trees called \defn{gene trees}, which represent the
%evolutionary history of a section of genetic material. Transmission trees, on
%the other hand, are highly analogous to \defn{species trees}, whose tips are
%species and internal nodes are common ancestors. This analogy derives from the
%functional similarity between transmission and allopatric speciation. Hence,
%the potential discordance between transmission trees and viral phylogenies is
%the similar to that between gene and species trees, which is called
%\defn{incomplete lineage sorting}.

%What we have just discussed is a two-step procedure for estimating the
%transmission tree. First, a viral phylogeny is constructed from genetic
%sequence data, and then it is rooted and time-scaled into a transmission tree.
%This approach is straightforward, frequently used, and has the advantage of
%leveraging tried-and-true tools for phylogenetic inference. However, it also
%has drawbacks, perhaps the most obvious being the multiplication of errors
%produced by the separate steps. One commonly used alternative method is to
%directly estimate a time-scaled phylogeny by simultaneously inferring the tree
%topology, its root and branch lengths, and the parameters of a \defn{molecular
%clock} model. A molecular clock is a hypothesis about the evolutionary rates
%along the branches of the tree, such as that they are all equal (a \defn{strict
%clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
%clock}). This inference is usually done in a Bayesian framework using
%\gls{MCMC}, so that prior information (including the tip dates) can be included
%in the analysis, and the so-called nuisance parameters of the molecular clock
%model can be marginalized out. Software packages for performing these analyses
%include BEAST~\autocite{bouckaert2014beast} and
%MrBayes~\autocite{ronquist2012mrbayes}.

%Other methods are tailor-made for inferring transmission trees.
%\textcite{didelot2014bayesian} develop a Bayesian approach which allows
%transmissions to occur anywhere along the
%branches of a transmission tree, rather than being constrained to the branching
%points in the viral phylogeny. The method requires sampling of every infected
%individual, although the authors indicate that it could be extended to relax
%this assumption. \textcite{cottam2008integrating} describe a likelihood-based
%method which enumerates all transmission trees consistent with an established
%phylogeny, assigning each a likelihood based on other epidemiological data.
%This approach is novel in its integration of data from multiple sources,
%however because it enumerates a large portion of the tree space, it is unlikely
%to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint
%likelihood function integrating temporal, geographic, and genetic observations,
%and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
%likelihood function. Their approach can handle missing data and produces high
%resolution transmission trees when multiple types of data are available. A
%different approach is undertaken by \textcite{jombart2011reconstructing}, who
%describe a method to build transmission trees directly from sequence data,
%contingent on the common ancestors also being sampled. This makes the method
%attractive for slow-evolving pathogens, but less practical for viral outbreaks
%where samples from common ancestors are unlikely to be available.

@@

To perform phylodynamic inference, we must be able to extract quantitative
information from viral phylogenies. What is informative about a phylogeny,
beyond the demographic characteristics of the individuals it relates, is its
\defn{shape}. The shape of a phylogeny has two components: the topology, and
the distribution of branch lengths~\autocite{mooers1997inferring}. Methods of
quantifying tree shape fall into two categories: summary statistics, and
pairwise measures. Summary statistics assign a numeric value to each individual
tree, while pairwise measures quantify the similarity between pairs of trees.

One of the most widely used tree summary statistics is Sackin's
index~\autocite{shao1990tree}, which measures the imbalance or asymmetry in a
rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
branches between that tip and the root. The unnormalized Sackin's index is
defined as the sum of all $N_i$. It is called unnormalized because it does not
account for the number of tips in the tree. Among two trees having the same
number of tips, the least-balanced tree will have the highest Sackin's index.
However, among two equally balanced trees, the larger tree will have a higher
Sackin's index. This makes it challenging to compare balances among trees of
different sizes. To correct this, \textcite{kirkpatrick1993searching} derive
the expected value of Sackin' index under the Yule
model~\autocite{yule1925mathematical}. Dividing by this expected value
normalizes Sackin's index, so that it can be used to compare trees of different
sizes. An example of a pairwise measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the absolute difference between the two plots is integrated between 0
and 1. In the context of infectious diseases, the \gls{ltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = \langle \varphi(x), \varphi(x')\rangle,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function $K$ is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. The feature space into which trees are mapped is the space of all
possible \defn{subset trees}, which are subtrees that do not necessarily extend
all the way to the tips. The subset-tree kernel was originally developed for
comparing parse trees in natural language processing~\autocite{collins2002new}
and did not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

@@

Several properties of scale-free networks are relevant in epidemiology.  The
high-degree hub nodes are known as
\defn{superspreaders}~\autocite{kemper1980identification}, which have been
postulated to contribute in varying degree to the spread of diseases such as
\gls{HIV}~\autocite{stadler2013uncovering} and
\gls{SARS}~\autocite{shen2004superspreading}. Scale-free networks have no
epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
arbitrarily low transmissibility can persist at low levels indefinitely. This
is in contrast with homogeneously mixed populations, in which transmissibility
below the epidemic threshold would result in exponential decay in the number of
infected individuals and eventual extinction of the
pathogen~\autocite{anderson1992infectious}.

@@

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2011contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by
\textcite{volz2009phylodynamics}. The concordance between the two skyrides, as
well as the relationship between the skyride and prevalence curve, was
qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and did find a
moderate correspondence between the two in some cases.

@@

\glsreset{SMC}

\subsection{Overview and notation}

\Gls{SMC} is the name for a family of statistical inference methods which rely
on approximating probability distributions of interest with large collections
of \defn{particles}, here denoted
$\set{x^{(k)}}$~\autocite{doucet2001introduction, liu2008monte}. These
collections or \defn{populations} are constructed to form a \defn{Monte Carlo
approximation} to some distribution of interest $\pi$, meaning that the
empirical distribution of the particles converges in distribution to $\pi$ as
the population size gets large~\autocite{liu2001theoretical}. The word
\defn{sequential} is used because the particle populations are modified in an
iterative fashion over time, for example, to incorporate new evidence.

@@

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. Note that a Markov kernels have nothing to
do with the kernel functions defined in \cref{subsec:treeshape}, other than
sharing a name (the word ``kernel'' is ubiquitous in mathematics).

\subsection{Sequential importance sampling for sequential Monte Carlo}
\label{subsec:sis}

\Gls{SIS} is one type of \gls{SMC} method, whose aim is to sample from a
distribution $\pi$ on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1,
\ldots, x_d)$. The basis of \gls{SIS} is \gls{IS}, which is a method of
estimating summary statistics of distributions which are known only up to a
normalizing constant, and therefore cannot be sampled from directly. That is,
if $\pi$ is such a distribution and $f$ is any real-valued function, \gls{IS}
is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We can
express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
we need to obtain an estimate of $\pi(f)$. Although this is a simple and
elegant approach, the drawback is that the variance of the estimate is
proportional to the variance of the importance weights~\autocite{liu2008monte},
which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
the practical use of \gls{IS} on its own is limited, since it depends on
finding an importance distribution similar to $\pi$, which we usually know very
little about \textit{a priori}.

The objective of \gls{SIS} is to build up an importance distribution $\eta$ for
$\pi$ sequentially. By the general product rule, $\pi(\vec{x})$ can be
decomposed as
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ can be written recursively as
\begin{align}
  \label{eq:sisw}
  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
\end{align}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. This choice is made on a
problem-specific basis, taking any available information about $\pi(x_i \mid
\vec{x_{i-1}})$ into account (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).  One potential
choice for $\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid
\vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
prior distribution may be used. The exact form of $\eta(x_i, \vec{x_{i-1}})$
which minimizes the variance of the weights is called the \defn{optimal
kernel}~\autocite{cappe2007overview}, the name deriving from the fact that
$k(x_i, \vec{x_{i-1}}) = \eta(x_i, \vec{x_{i-1}})$ is a Markov kernel. In some
applications, it is possible to approximate the optimal kernel or even compute
it explicitly.

@@

\subsection{The sequential Monte Carlo sampler}
\label{subsec:smcsamp}

@@

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $x$ and outcomes $y$. A
\defn{parametric} model defines a family of possible relationships between data
and outcomes, indexed by one or more numeric parameters $\theta$. A
\defn{statistical} model describes the relationship between data and outcomes
in terms of probabilities. Statistical models define, either explicitly or
implicitly, the probability of observing $y$ given $\vec{x}$ and, if the
model is parametric, $\theta$. Note that it is entirely possible to have no
data $\vec{x}$, only observed outcomes $y$. In this case, a model would
describe the process by which $y$ is generated.
